---
title: "Penguin Analysis"
author: "Alberto F Cabrera"
format:
  html:
    code-link: true
    code-fold: true
    code-summary: "Show the code"
    self-contained: true
    embed-resources: true
    toc: true
    theme: Zephyr
    number-sections: false
execute: 
  warning: false
  message: false
  echo: false
editor_options: 
  chunk_output_type: console
fig-cap-location: margin
---

## Introduction

This is our first quarto document.

```{r loading_files}
#| echo: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
library(janitor)
library(readxl)
```

## Loading penguin data

Loading penguin data and summarizing it. The dataset comes from [Palmerpenguins](https://allisonhorst.github.io/palmerpenguins/). It describes the *physical* characteristics of three **penguin** ***species***.

```{r loading_data_summarizing}
#| echo: true
#| warning: false

# Read the clean dataset

#Option offered by the author
#data <- readRDS(file = "../data/data_clean.rds")

data <- readRDS("~/Documents/R-books/productive-r-workflow/data/data_clean.rds")


#Sumarizing data

summary(data)

```

Calculating mean bill length for different species and islands

```{r calculating_mean}
#| echo: true


# Calculating mean bill length for different species and islands using dplyr
data |> 
  filter(species == "Adelie") |> 
  group_by(island)  |> 
  summarize(mean_bill_length = round(mean(bill_length_mm, na.rm = TRUE), 2))

```

```{r plotting_penguin_data_detailed}
#| echo: true
#| fig-align: "center"
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Relationship between bill <b>length</b> and bill <b>depth</b>. <b>All</b> data points included."

data |> 
  ggplot(
    aes(x = bill_length_mm, y = bill_depth_mm)
  ) +
    geom_point(color = "#69b3a2") +
    labs(
      x = "Bill Length (mm)",
      y = "Bill Depth (mm)",
      title = paste("Surprising relationship?")
    ) + 
  theme_minimal()

```

It is also interesting to note that `bill length` a and `bill depth` are quite different from one specie to another. This is summarized in the 2 tables below:

```{r two_column_layout}
#| layout-ncol: 2

data %>%
 group_by(species) %>% 
  summarise(average_bill_length = mean(bill_length_mm, na.rm = TRUE))

data %>%
 group_by(species) %>% 
  summarise(average_bill_depth = mean(bill_depth_mm, na.rm = TRUE))

```

## Invoking function & performing plots

Loading the create-scatterplot function.

```{r loading_data}
#| echo: true

#Invoking functions file

#This option works if rending the quarto document
source(file = "functions.R")

#This option works if rendering the quarto document
#source(file = "../R/functions.R")

#This option works if ran individually
#source(file = "R/functions.R")

# Use the function in functions.R
#create_scatterplot(data, "Adelie", "Torgersen")

```

### Scatter plot options

Scatter plot with width = 3

```{r scatter_plot_width_3_inches}
#| echo: false
#| fig-width: 3

create_scatterplot(data,"Adelie","Torgesen")


```

Scatter plot with width = 6

```{r scatter_plot_width_6_inches}
#| echo: false
#| fig-width: 6

create_scatterplot(data,"Adelie","Torgesen")


```

Scatter plot with width = 9

```{r scatter_plot_width_9_inches}
#| echo: false
#| fig-width: 9

create_scatterplot(data,"Adelie","Torgesen")
```

Aligning the figure of 6 inches

Aligning to the center

```{r scatter_plot_width_6_inches_center}
#| echo: false
#| fig-width: 6
#| fig-align: center

create_scatterplot(data,"Adelie","Torgesen")

```

Aligning to the left

```{r scatter_plot_width_6_inches_left}
#| echo: false
#| fig-width: 6
#| fig-align: left

create_scatterplot(data,"Adelie","Torgesen")

```

Aligning to the right

```{r scatter_plot_width_6_right}
#| echo: true
#| fig-width: 6

create_scatterplot(data,"Adelie","Torgesen")


create_scatterplot(data,"Gentoo","Biscoe")

```

Figure capture margins

```{r capture_margins}
#| echo: true

create_scatterplot(data,"Adelie","Torgesen")


create_scatterplot(data,"Gentoo","Biscoe")



```

Adding equations:

This formula $f(x) = x^2$ is an example

The $X^2 =$ is

Adding a kable() report

```{r kable_penguin_report}
#| echo: true

library(tidyverse)
library(knitr)
library(palmerpenguins)

table_penguins <- palmerpenguins::penguins %>%
 group_by(species) %>% 
  summarise(average_bill_depth = mean(bill_depth_mm, na.rm = TRUE),
            average_bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE))


kable(table_penguins)


```

```{r author_approach}
#| echo: true
library(tidyverse)
library(knitr)
library(palmerpenguins)



bill_length_per_specie <- palmerpenguins::penguins |> 
 group_by(species)  |>  
  summarise(average_bill_length = mean(bill_length_mm, na.rm = TRUE))

kable(bill_length_per_specie)

bill_depth_per_specie <- palmerpenguins::penguins %>%
 group_by(species) %>% 
  summarise(average_bill_depth = mean(bill_depth_mm, na.rm = TRUE))

kable(bill_depth_per_specie)

bill_length_adelie <- bill_length_per_specie %>%
  filter(species == "Adelie") %>%
  pull(average_bill_length) %>%
  round(2)


```

For instance, the average bill length for the specie `Adelie` is `r bill_length_adelie`.

Now, let’s check the relationship between bill depth and bill length for the specie `Adelie` on the island `Torgersen`:

```{r}
#Calling function

# Source functions
source(file = "functions.R")

```

```{r relationship_species}
#| echo: true

# Source functions
#source(file = "functions.R")

p1 <- create_scatterplot(data, "Adelie", "Torgersen")
p2 <- create_scatterplot(data, "Chinstrap", "Dream")
p3 <- create_scatterplot(data, "Gentoo", "Biscoe")

p1 + p2 + p3


```
